---
title: "Ambient air quality, WHO--  EDA"
output: html_document
date: "2025-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(dplyr)
```

some things could explore-

- which cities have highest average AQI
- avaerage level of year pollutant per year
-city /  country with cleanest air on average - could look at general trend here, are they more/less developed?- this knowledge of the countries would this be considered expert knowledge


looking at population specifically.
how does air quality change as population increases,are different pollutant different in different pop sizes.


lets look at the dataset and clean the data set for each column: pm10_concentration, pm25_concentration, no2_concentration, population.
```{r}
dataset<- who_data
dataset

dataset[dataset == "NA"] <- NA


clean_data_pm10 <- dataset[!is.na(dataset$pm10_concentration), ]
clean_data_pm25 <- dataset[!is.na(dataset$pm25_concentration), ]
clean_data_no2 <- dataset[!is.na(dataset$no2_concentration), ]
clean_data_pop <- dataset[!is.na(dataset$population), ]

clean_data_all <- dataset[
  !is.na(dataset$population) &
  !is.na(dataset$no2_concentration) &
  !is.na(dataset$pm25_concentration) &
  !is.na(dataset$pm10_concentration),
]
clean_data_pm10 <- clean_data_pm10 %>%
  mutate(across(c(no2_concentration, pm25_concentration, pm10_concentration, population),
                as.numeric))
clean_data_pm25 <- clean_data_pm25 %>%
  mutate(across(c(no2_concentration, pm25_concentration, pm10_concentration, population),
                as.numeric))


clean_data_no2 <- clean_data_no2 %>%
  mutate(across(c(no2_concentration, pm25_concentration, pm10_concentration, population),
                as.numeric))

clean_data_pop <- clean_data_pop %>%
  mutate(across(c(no2_concentration, pm25_concentration, pm10_concentration, population),
                as.numeric))
clean_data_all <- clean_data_all %>%
  mutate(across(c(no2_concentration, pm25_concentration, pm10_concentration, population),
                as.numeric))

```

 lets calculate some average values of columns
 
```{r}
mean(clean_data_pm10$pm10_concentration)
mean(clean_data_pm25$pm25_concentration)
mean(clean_data_no2$no2_concentration)
mean(clean_data_pop$population)

```


```{r}


```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

